---
title: "Empirical Null Worlds"
author: "Myla Burton"
date: "2024-12-18"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Administrative stuff: getting the dataset and installing necessary packages

```{r load-data, message=FALSE, warning=FALSE}
# The URL to the raw CSV file on my GitHub repository.
url <- "https://raw.githubusercontent.com/mylaburton/null-worlds/refs/heads/main/ExtendedTS.csv"

# Read the CSV file.
ExtendedTS <- read.csv(url)

# Display the first few rows of the dataset to ensure all is working properly.
head(ExtendedTS)
```

# Testing for Stationarity in the ExtendedTS series 
## Augmented Dickey-Fuller (ADF) Test: the null is that the series has a unit root (i.e., the series is nonstationary)
```{r}
# Install and load the tseries package
install.packages("tseries")
library(tseries)

# Run the ADF Test for each of the state's series 
adf.test(ExtendedTS$Armenia)
adf.test(ExtendedTS$Azerbaijan)
adf.test(ExtendedTS$Belarus)
adf.test(ExtendedTS$Kazakhstan)
adf.test(ExtendedTS$Kyrgyzstan)
adf.test(ExtendedTS$Latvia)
adf.test(ExtendedTS$Lithuania)
adf.test(ExtendedTS$Russia)
adf.test(ExtendedTS$Ukraine)
```

## Kwiatkowski-Phillips-Schmidt-Shin Test: the null is that the series is stationary around a deterministic trend (trend stationarity or level stationarity). This is the opposite of the ADF.
```{r}
# Run the KPSS test for each of the state's series
kpss.test(ExtendedTS$Armenia)
kpss.test(ExtendedTS$Azerbaijan)
kpss.test(ExtendedTS$Belarus)
kpss.test(ExtendedTS$Kazakhstan)
kpss.test(ExtendedTS$Kyrgyzstan)
kpss.test(ExtendedTS$Latvia)
kpss.test(ExtendedTS$Lithuania)
kpss.test(ExtendedTS$Russia)
kpss.test(ExtendedTS$Ukraine)
```

## Phillips-Perron Test: the null is that the series has a unit root (i.e., the series is nonstationary)
```{r}
# Run the Phillips-Perron test for each of the state's series
pp.test(ExtendedTS$Armenia)
pp.test(ExtendedTS$Azerbaijan)
pp.test(ExtendedTS$Belarus)
pp.test(ExtendedTS$Kazakhstan)
pp.test(ExtendedTS$Kyrgyzstan)
pp.test(ExtendedTS$Latvia)
pp.test(ExtendedTS$Lithuania)
pp.test(ExtendedTS$Russia)
pp.test(ExtendedTS$Ukraine)
```

# Defining the Simulation Function 
```{r}
install.packages("urca")
library(urca)

simulate_null_worlds <- function(ExtendedTS, n_simulations = 10000, null_type = "russia_ignores", block_size = 5, seed = 123) {
  set.seed(seed)
  n <- nrow(ExtendedTS)
  
  # Creating a data frame to store the simulation results. We are storing the   estimated rank of cointegration for each simulation into this data frame.
  results <- data.frame(simulation = seq_len(n_simulations), cointegrated_rank = integer(n_simulations), cointegrating_vectors = I(vector("list", n_simulations)))
  
  # Counter for false positives
  false_positives <- 0
  
  # Starting the simulation loop. It will loop over 10,000 (the number of     simulations specified in n_simulations_)
  for(i in seq_len(n_simulations)) {
    permuted_data <- ExtendedTS
    
    # Applying the null hypothesis. Here we are simulating a null world by     shuffling the data under each of the   specified null hypotheses.
    if (null_type == "russia_ignores") {
    permuted_data$Russia <- sample(ExtendedTS$Russia) #Shuffling only Russia
    } else if (null_type == "everyone_ignores_each_other") {
      permuted_data <- as.data.frame(apply(ExtendedTS, 2, sample)) # Shuffling all series 
    } else if (null_type == "block_permutation") {
      permuted_data <- as.data.frame(matrix(nrow=n, ncol = ncol(ExtendedTS)))
      colnames(permuted_data) <- colnames(ExtendedTS)
      
      for (col in 1:ncol(ExtendedTS)) {
        series <- ExtendedTS[, col]
        num_blocks <- ceiling(n / block_size)
        blocks <- split(series, ceiling(seq_along(series) /block_size))
        shuffled_blocks <- sample(blocks)
        permuted_series <- unlist(shuffled_blocks)[1:n]
        permuted_data[, col] <- permuted_series
  }
    } else {
    # Stopping if an invalid null_type is passed
    stop("Invalid null_type")
    }
    
    # Running the Johansen test
    johansen_test <- ca.jo(permuted_data, type = "trace", ecdet = "const", K = 2)
    trace_stat <- johansen_test@teststat
    critical_values <- johansen_test@cval
    cointegrated_rank <- min(sum(trace_stat > critical_values[, 2]), ncol(permuted_data) - 1)
    
    # Storing the results
    results$cointegrated_rank[i] <- cointegrated_rank
    
    # Extracting and storing the cointegrating vectors if the rank is above 0
    if (cointegrated_rank > 0) {
      results$cointegrating_vectors[[i]] <- johansen_test@V[, 1:cointegrated_rank]
    } else {
      results$cointegrating_vectors[[i]] <- NA # No cointegration detected
    }
    
    # Check if the permuted rank exceeds the observed rank
    if (cointegrated_rank >= observed_rank) {
      false_positives <- false_positives + 1
    }
  }
  # Calculate false positive rate
  false_positive_rate <- false_positives / n_simulations
  
  # Create a summary table of cointegrated ranks
  rank_table <- table(results$cointegrated_rank)
  
  # Debugging false positive rate
  cat("False positives:", false_positives, "False positive rate:", false_positive_rate, "\n")
  
  return(list(results = results, false_positive_rate = false_positive_rate))
  return(results)
}
```

# Observed Cointegration Test and Null World Simulations 
```{r}
# Preparing the data with the relevant series
combined_data <- ExtendedTS[, c("Armenia", "Azerbaijan", "Belarus", "Ukraine", "LitLat", "Kazakhstan", "Kyrgyzstan", "Russia")]

# Computing the observed rank for the original dataset using the Johansens Cointegration Test
observed_test <- ca.jo(combined_data, 
                       type = "trace",
                       ecdet = "const",
                       K = 2)
observed_trace_stat <- observed_test@teststat
observed_critical_values <- observed_test@cval
observed_rank <- min(sum(observed_trace_stat > observed_critical_values[,2]),
                     ncol(combined_data)-1)

# Null World 1: Everybody Ignores Everybody
results_null1 <- simulate_null_worlds(combined_data, n_simulations = 10000, null_type = "everyone_ignores_each_other")

# Null World 2: Russia Ignores Everybody
results_null2 <- simulate_null_worlds(combined_data, n_simulations = 10000, null_type = "russia_ignores")

# Null World 3: Block Permutations
results_null3 <- simulate_null_worlds(combined_data, n_simulations = 10000, null_type = "block_permutation", block_size = 5)

# Summarizing the results
cat("Observed Cointegrated Rank:", observed_rank, "\n")
cat("Null World 1 (Everbody Ignores Everbody):\n")
cat("False Positive Rate for Null 1:", results_null1$false_positive_rate, "\n\n")

cat("Null World 2 (Russia Ignores Everyone):\n")
cat("False Positive Rate for Null 2:", results_null2$false_positive_rate, "\n\n")

cat("Null World 3 (Block Permutations):\n")
cat("False Positive Rate for Null 3:", results_null3$false_positive_rate, "\n\n")

```

# Creating the Visual Plots of the Null World Distributions
```{r}
install.packages("ggplot2")
library(ggplot2)

# A function to create and display the plot for each of the null world's distributions
create_plot <- function(results, observed_rank, null_world_name) {
  # Creating a data frame for plotting
  plot_data <- data.frame(CointegratedRank = results$results$cointegrated_rank)
  
  # Generating the plot
  ggplot(plot_data, aes(x = CointegratedRank)) +
    geom_histogram(bins = 30, fill = "skyblue", alpha = 0.7, boundary = 0) + geom_vline(aes(xintercept = observed_rank), color = "red", linetype = "dashed", linewidth = 1) +
    labs(
      title = paste("Null Distribution for", null_world_name),
      x = "Cointegrated Rank",
      y = "Frequency",
      subtitle = "Red dashed line indicates the observed cointegrated rank"
    ) +
    theme_minimal()
}

# Generating and printing the plots for each null world

# Plot for Null World 2: Russia Ignores Everyone
plot2 <- create_plot(results_null2, observed_rank, "Null World 2 (Russia Ignores Everyone)")
print(plot2)

# Plot for Null World 1: Everyone Ignores Each Other
plot1 <- create_plot(results_null1, observed_rank, "Null World 1 (Everyone Ignores Each Other)")
print(plot1)

# Plot for Null World 3: Block Permutation
plot3 <- create_plot(results_null3, observed_rank, "Null World 3 (Block Permutation)")
print(plot3)
```

